Binning Strategies and Related Loss for Large Data

### 2015-02-28 Paper.Rnw
Continued (2) work on binning rewrite

\ktm{  Paragraph: Discussion of types of deterministic rectangular binning.  Standard and Quantile. Table~\ref{tab:rectbinning} }

\begin{table}[hbtp]
\centering
\begin{tabular}{lll} \hline
 & Bin Boundaries & Bin Centers \\ 
 \hline  
General &  $ \{\beta_j \text{ }|\text{ } \beta_j > \beta_{j-1} \} $ & $\{x_j^\ast \text{ }|\text{ } x_j^\ast = (\beta_{j-1}+ \beta_j)/2 \}$ \\
%-------------
Standard \hspace{0.5cm} & $ \{\beta_j \text{ }|\text{ } \beta_j = \beta_{j-1} + \omega_X \} $\hspace{0.5cm} & $\{x_j^\ast \text{ }|\text{ } x_j^\ast = \beta_{j-1} + \omega_X/2 \}$ \\
%-------------
Quantile & $ \{\beta_j \text{ }|\text{ } \beta_j = Q_X(j/J) \} $  & $\{x_j^\ast \text{ }|\text{ } x_j^\ast = Q_X((j-0.5)/J) \}$ \\
%-------------
Random &  ---  & $\{x_j^\ast \text{ }|\text{ } x_j^\ast > x_{j-1}^\ast \}$  \\
\hline
\end{tabular}
\caption{Rectangular Binning Specifications}
\label{tab:rectbinning}
\end{table}

\ktm{ An alternative to the rectangular binning processes discussed above, is the random binning algorithm which utilizes a non-deterministic bin function $b^r_X(\cdot)$ which randomly assigns a value $x_i$ to a bin center, $x^\ast$, from a set of possible bins. In this paper, we will consider the simplest case of just two bins, so that without loss of generality we can assume that $x_i$ lies between bin centers $x^\ast_j$ and $x^\ast_{j+1}$. The bin function assigns $x_i$ to a bin center with a probability depending on the distance to that bin center; the closer a value is to a bin center, the higher the probability the value is assigned to that bin center. More formally, }
%
\ktm{ \begin{eqnarray}\label{randbin1}
b^r_X(x_i) = \left\{\begin{array}{ll} 
  x^\ast_j & \text{ with probability } (x^\ast_{j+1} - x_i)/(x^\ast_{j+1} -x^\ast_{j}) \\
  x^\ast_{j+1} &\text{ with probability }  (x_i - x^\ast_{j})/(x^\ast_{j+1} -x^\ast_{j})
  \end{array}\right.
\end{eqnarray}  
}
%
\ktm{  for $x_i \in [x^\ast_{j+1}, x^\ast_{j}]$.  This probability structure is inversely proportional to the distance from each bin center. This method is easily extensible to also map $x_i$ into more than two bins and can accommodate non-uniform bin sizes. }

>>>>

\ktm{ \textit{Standard rectangular binning} is a special cases of this general rectangular binning that uses intervals of equal size for all bins; thus only the origin of the first bin, $\beta_0$, and a binwidth, $\omega_X$, need to be specified. Standard rectangular binning is necessarily used in the construction of all histograms and is ideal for visualizing density because the consistency of binwidth makes the display of frequency proportional to density. }

\ktm{  \textit{Quantile binning} is another option that divides the range of the observations into bins each containing equal number of points. The $j^{th}$ bin intervals is $(Q_X((j-1)/J),Q_X((j)/J)]$, where $Q_X(p)$ is the the $p^{th}$ empirical quantile using the inverse empirical distribution function. Note that this binning approach is \textit{not} desirable for visualizing density patterns, as it effectively balances the frequency counts in all bins; it does however have desirable properties for creating discernable groups in visualization using discrete color palettes. }

\ktm{  The bin boundaries and centers for each type of rectangular binning algorithm discussed above can be found in Table~\ref{tab:rectbinning}.  }

\begin{table}[hbtp]
\centering
\begin{tabular}{lll} \hline
 & Bin Boundaries & Bin Centers \\ 
 \hline  
General &  $ \{\beta_j \text{ }|\text{ } \beta_j > \beta_{j-1} \} $ & $\{x_j^\ast \text{ }|\text{ } x_j^\ast = (\beta_{j-1}+ \beta_j)/2 \}$ \\
%-------------
Standard \hspace{0.5cm} & $ \{\beta_j \text{ }|\text{ } \beta_j = \beta_{j-1} + \omega_X \} $\hspace{0.5cm} & $\{x_j^\ast \text{ }|\text{ } x_j^\ast = \beta_{j-1} + \omega_X/2 \}$ \\
%-------------
Quantile & $ \{\beta_j \text{ }|\text{ } \beta_j = Q_X(j/J) \} $  & $\{x_j^\ast \text{ }|\text{ } x_j^\ast = Q_X((j-0.5)/J) \}$ \\
%-------------
Random &  ---  & $\{x_j^\ast \text{ }|\text{ } x_j^\ast > x_{j-1}^\ast \}$  \\
\hline
\end{tabular}
\caption{Rectangular and Random Binning Specifications}
\label{tab:rectbinning}
\end{table}

\ktm{ An alternative to the rectangular binning processes discussed above, is the random binning algorithm which utilizes a non-deterministic bin function $b^r_X(\cdot)$ which randomly assigns a value $x_i$ to a bin center, $x^\ast$, from a set of possible bins. In this paper, we will consider the simplest case of just two bins, so that without loss of generality we can assume that $x_i$ lies between bin centers $x^\ast_j$ and $x^\ast_{j+1}$. The bin function assigns $x_i$ to a bin center with a probability depending on the distance to that bin center; the closer a value is to a bin center, the higher the probability the value is assigned to that bin center. More formally, }
%
\ktm{ \begin{eqnarray}\label{randbin1}
b^r_X(x_i) = \left\{\begin{array}{ll} 
  x^\ast_j & \text{ with probability } (x^\ast_{j+1} - x_i)/(x^\ast_{j+1} -x^\ast_{j}) \\
  x^\ast_{j+1} &\text{ with probability }  (x_i - x^\ast_{j})/(x^\ast_{j+1} -x^\ast_{j})
  \end{array}\right.
\end{eqnarray}  
}
%
\ktm{  for $x_i \in [x^\ast_{j+1}, x^\ast_{j}]$.  This probability structure is inversely proportional to the distance from each bin center.  In Table~\ref{tab:rectbinning} we note that this random binning algorithm does not specify bin boundaries; only a sequence of bin centers.  This method is easily extensible to also map $x_i$ into more than two bins and can accommodate non-uniform distribution of bin centers.  }
